---
title: "SST Calibration with AIMS temperature logger data"
author: "Catherine Kim, John Lyons, Paula Hatum"
date: "2023-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal of this notebook

This notebook will demonstrate how to access the RIMReP `geoparquet` collection for AIMS Sea Surface Temperature Monitoring Program. This dataset includes sea temperature data since 1980 for tropical and subtropical coral reefs around Australia, including sites at the Great Barrier Reef.

We will also extract the coordinates for all sites sampled in this monitoring program. This way we can extract data for the site of our choice using the name of the site of our interest, without the need to know their coordinates.

# Loading libraries

```{r libraries, results = "hide", message = F, warning = F}
#Accessing S3 bucket
library(arrow)
#Data manipulation
library(dplyr)
library(magrittr)
library(stringr)
#Managing dates
library(lubridate)
#Plotting
library(ggplot2)
library(rnaturalearth)
#Managing spatial data
library(sf)
library(wkb)
```

# Connecting to RIMReP collection

As mentioned above, we will connect to the AIMS Sea Surface Temperature Monitoring Program. This can take a minute or so.

```{r connect}
#Establishing connection
data_bucket <- s3_bucket("s3://rimrep-data-public/091-aims-sst/test-50-64-spatialpart/")

#Accessing dataset
data_df <- open_dataset(data_bucket)
```

## Exploring dataset structure

We can explore the dataset `schema` to understand its structure, and identify the fields that are relevant to us.

```{r explore}
#Checking data structure
data_df$schema
```
We can see that there are a number of variables available in this dataset. We will need to access three variables to create a list of all sampled sites in this monitoring program:  
- `site`, which includes the name of all sites where temperature loggers were deployed  
- `subsite`, which includes a shorten version of the site name and an indication of where loggers where deployed (see below for more details)  
- `geometry`, which includes latitude and longitude coordinates in [well-known binary (WKB)](https://loc.gov/preservation/digital/formats/fdd/fdd000549.shtml) format  
  
Based on the `subsite` information, we will create a new column that we will call `deployment_location`. If `subsite` string includes `SL`, this indicates the temperature logger was deployed on a reef slope (depths of 3 m or more). If this is the case, our new column will be labelled `reef slope`. If `FL` is included in the `subsite` string, this means the temperature logger was deployed on a reef flat (areas less than 3 m deep). In this case, we will classify this as `reef flat` in our new column.  
  
Next, we will transform the information contained in the `geometry` field into coordinate pairs (i.e., latitude and longitude in decimal degrees).  
  
**Note** that you can get additional information about the AIMS dataset by exploring its metadata as shown below.

```{r metadata}
#Checking metadata
data_df$metadata
```
# Extracting variables of our interest from dataset

We can extract location (coordinates) from the AIMS dataset by using `dplyr` verbs as shown below.

```{r sites_bin}
sites <- data_df %>% 
  #We select unique sites included in the dataset
  distinct(site, subsite, geometry) %>%
  #This will load them into memory
  collect()

#Checking results
glimpse(sites)
```

## Creating `deployment_location` column

We will use the `subsite` column to create the categories in our new column, as explained in the [exploring the dataset structure](#exploring-dataset-structure). If no condition is met, then we will label the row as *other*.

```{r dep_loc}
sites <- sites %>% 
  #Adding new column - Given categories based on a condition
  mutate(deployment_location =  case_when(str_detect(subsite, "FL[0-9]{1}") ~ "reef flat",
                                          str_detect(subsite, "SL[0-9]{1}") ~ "reef slope",
                                          #If no condition is met, return other
                                          T ~ "other"))

#Checking results
glimpse(sites)
```

## Extracting sites and coordinates from dataset 
### Transforming `geometry` format
```{r sites_deg}
sites <- sites %>% 
  #Adding column with coordinate pairs in degrees
  mutate(coords_deg = readWKB(geometry) %>% st_as_sf()) %>% 
  #Separating coordinate pairs into latitude and longitude columns
  mutate(lon = st_coordinates(coords_deg)[,"X"],
         lat = st_coordinates(coords_deg)[,"Y"])

#Checking results - We will exclude the geometry column
sites %>% 
  select(!geometry) %>% 
  arrange(site) %>% 
  head()
```


# Plotting map of sampled sites in this dataset

We will make a map of all sites included in the AIMS Water Temperature Monitoring Program. We will also highlight the location of two sites: **Hayman Island** and **Heron Island** in red. We will use these highlighted sites as examples to show you how to extract data at multiple sites of interest.

```{r map_sites}
#Loading basemap with countries in Oceania and Asia
oce_asia <- ne_countries(continent = c("oceania", "asia"), 
                                     returnclass = "sf")

#Creating map with all sites included in dataset
#First we plot the basemap with the countries in Oceania and Asia
oce_asia %>% 
  ggplot()+
  geom_sf()+
  #Plotting monitoring sites
  geom_point(inherit.aes = F, data = sites, aes(x = lon, y = lat))+
  geom_point(inherit.aes = F, data = sites[sites["site"] == "Hayman Island" | sites["site"] == "Heron Island",], 
             aes(x = lon, y = lat), color = "red", size = 2)+
  #Changing map limits - Adding a buffer of 2 degree around max and min site coordinates
  lims(y = c(floor(min(sites$lat))-2, ceiling(max(sites$lat))+2),
       x = c(floor(min(sites$lon))-2, ceiling(max(sites$lon))+2))+
  #Removing default grey background
  theme_bw()
```

# Extracting data for sites of interest - Heron Island and Hayman Island

We can use the information in our `sites` variable to extract data for a site of interest. All we need to know is the name of the site, as shown below.  
  
Here, we will use data collected at **Hayman Island** and **Heron Island** as an example. We will calculate monthly means at each site, so we can plot time series that will show us how temperature has changed over time.

```{r site_coords}
sites_coords <- sites %>% 
  #Using random site as example
  filter(site %in% c("Hayman Island", "Heron Island")) %>% 
  #Extracting latitude and longitude coordinates
  select(site, lon, lat, deployment_location)

#Checking results
sites_coords
```

We can see that there are multiple coordinates for each site, but they are located in close proximity to one another. In this case, both sites include instrument deployments in the reef flat and slope. We will use this information to extract data from the original AIMS dataset.  
  
Note that `qc_val` is the variable containing quality-controlled temperature data in this dataset.

```{r extract}
sites_temp <- data_df %>% 
  #We will only keep data for the sites of our interest
  inner_join(sites_coords, by = c("site", "lon", "lat")) %>% 
  #Turning results into data frame
  collect()
```

We will now calculate the monthly temperature means at each site and deployment location (i.e., reef flat, reef slope, other).

```{r means}
sites_mmtemp <- sites_temp %>% 
  #We will two new columns: year and month to calculate monthly means
  mutate(year = year(time),
         month = month(time)) %>%
  #We will now group data by month, year and site
  group_by(year, month, site, deployment_location, lon, lat) %>%
  #Calculating monthly means for temperature at each site
  summarise(temp_monthly_mean = round(mean(qc_val, na.rm = TRUE), 2),
            #Total number of observations used in monthly mean calculations
            tot_obs = n()) %>%
  #Arranging in chronological order
  arrange(site, deployment_location, year, month)

#Checking results
head(sites_mmtemp)
```

## Plotting timeseries

We can now use `ggplot2` to create a plot showing how temperature has changed over time. We will save the plot as a variable so we can save it to our local machine later.

```{r timeseries_plot}
temp_plot <- sites_mmtemp %>% 
  #Combining year and month columns into one
  mutate(date = ym(paste0(year, "-", month))) %>% 
  #Plotting temperature of y axis and time on x axis. Color data by site.
  ggplot(aes(x = date, y = temp_monthly_mean, color = site))+
  #Plot data as points and lines
  geom_point()+
  geom_line()+
  #Creating subplots for each site for easy comparison
  facet_grid(~deployment_location)+
  #Removing default grey background
  theme_bw()+
  #Change position of legend
  theme(legend.position = "top", axis.title.x = element_blank())+
  #Change position and title of legend
  guides(colour = guide_legend(title.position = "top", title.hjust = 0.5, title = "Site names"))+
  labs(y = expression("Mean monthly temperature  " ( degree~C)))

#Checking plot
temp_plot
```

## Read in sea surface temperature (SST) from NOAA Extended Reconstructed SST



